 Casting multiple shadows: high-dimensional interactive data visualisation with tours and embeddings

### 2020-03-24 paper.Rmd
wording
rejigging sections
update paper

---
title: " Casting multiple shadows: interactive data visualisation with tours and embeddings"
author:
  - name: Stuart Lee
  - name: Di Cook
date: "`r Sys.Date()`"
output: bookdown::html_document2
bibliography: liminal.bib
---

@@

<!--
- review of embedding algorithms, interactive graphics, tours
- technology
- design
- examples/benchmarks
  - Kevin moon
  - pdfsense
  - single cell
  - quick draw

https://jlmelville.github.io/smallvis/init.html#other_initialization_options

-->

# Introduction

High dimensional data is increasingly prevalent in the natural sciences and
beyond but presents a challenge to the analyst in terms of both data cleaning /
preprocessing and visualisation. Methods to embed data from a high-dimesional
space into a low-dimensional one now form a core step of the data analysis
workflow where they are used to ascertain hidden structure and denoise data for
downstream analysis (thereby nullifying the 'curse of dimensionality').

@@

Non-linear methods generally perform preprocessing on the high-dimensional data
such as generating a neighborhood graph and perform transformations on the
preprocessed form. We restrict our attention to three methods that are commonly
used in high-throughput biology: t-distributed stochastic neighbor embedding
(t-SNE), uniform manifold alignment and projection (UMAP), and potential of
heat-diffusion for affinity-based transition embedding (PHATE). The t-SNE
algorithm estimates the similarity of (Euclidean) distances of points in a high
dimensional space using a Gaussian distribution and then estimates a
configuration in the low dimensional embedding space by modelling similarities
using a t-distribution with 1 degree of freedom. The resulting configuration is
the one that minimizes the Kulback-Leibler divergence between the two
distributions. A recent theoretical contribution by @Linderman2019-dq proved
that t-SNE can recover spherical and well separated cluster shapes, and
proposed new approaches for tuning the optimisation parameters. It is a known
problem that t-SNE can have trouble recovering global structure and that
configurations can be highly dependent on how the algorithm is initialised and
parameterised [@Wattenberg2016-ji; @Kobak2019-lm]. UMAP is a method that is
related to LargeVis [@Tang2016-oz], and like t-SNE acts on the k-nearest
neighbor graph. Its main differences are that it uses a different cost function
(cross entropy) which is optimized using stochastic gradient descent and
defines a different kernel for similiarites in the low dimensional space. Due
to it's computational speed it's possible to generate UMAP embeddings in more
than three dimensions. Finally, PHATE, inspired by diffusion maps, is based on
estimating an affinity matrix via a distance matrix and k-nearest neighbors
graph. The algorithm denoises estimated distances in high dimensional space via
transforming the affinity matrix into a Markov transition probaility matrix and
diffusing this matrix over a fixed number of time steps. Then the diffused
probabilities are transformed once more to construct a distance matrix, and
multidimensional scaling is performed to construct a 2d embedding for
visualization.

As part of a visualization workflow, it's important to consider the perception
and interpretation of embedding methods as well. @Sedlmair2013-pn showed that
2D scatter plots were mostly suffucient for detecting class seperation, however
they noted that often multiple embeddings were required. For the task of
cluster identification, @Lewis2012-ai showed experimentally that novice users
of non-linear embedding techniques were more likely to consider clusters of
points on a 2d scatter plot to be the result of a spurious embedding compared
to advanced users who were aware of the inner workings of the embedding
algorithm.

A complimentary approach for visualizing structure in high dimensional data is
the tour. A tour is a sequence of projections of a high dimensional dataset
onto a low-dimensional orthonormal basis matrix, that is represented as a
dynamic visualization. The sequence of generated bases are interpolated to form
the tour path, allowing a user to explore the subspace of projections. A grand
tour corresponds to choosing new bases at random, and can give an overview of
the structure in the data. Instead of picking projections at random, a guided
tour can be used to generate a sequence 'interesting' projections as quantified
by an index function. Given the dynamic nature of the tour, user interaction is
important for controlling and exploring the visualisation: the tour has been
used previously by @Wickham2015-cx as tool for exploring statistical model fits
and by @Buja1996-fk for exploring factorial experimental designs.

While there has been much work on the algorithmic details of the aformentioned
embedding methods, there has been relatively few tools designed to assist users
to interact with these techniques and assist them in making comparsions between
embeddings and performing the aforementioned cluster orientation tasks. Several
interactive interfaces have been proposed for evaluating or using embedding
techniques: the Sleepwalk interface provides a click and highlight
visualisation for colouring points in an embedding according to their distance
in the original high-dimensional space [@Ovchinnikova2019-gf]. The work by
@Pezzotti2017-cz provides a user guided and modified form of the t-SNE
algorithm, that allows users to modified optimisation parameters in real-time.
Similarly, the embedding projector is a web interface to running UMAP, t-SNE or
PCA live in the browser and provides interactions to color points, and
highlights nearest neighbors [@Smilkov2016-hp].

There is no one-size fits all: finding an appropriate embedding for a given
dataset is a difficult and somewhat poorly defined problem. For non-linear
methods, there are a lot of parameters to explore that can have an effeft on
the resulting visualisation and interpretation. Interfaces for evaluating
embeddings require interaction but should also be able to be incorporated into
an analysts workflow. We propose a more pragmatic workflow inspired by
incorporating interactive graphics and tours with embeddings that allows users
to see a global overview of their high dimensional data and assists them with
cluster orientation tasks. This workflow is incorporated into an R package
called `liminal` (Available: https://github.com/sa-lee/liminal).

# Design

@@

## Finding Gestalt: focus and context

To investigate latent structure and the shape of a high dimensional dataset, a
tour can be run without the use of an external embedding. It is often useful to
first run principal components on the input as an intial dimension reduction
step, and then tour a subset of those components instead, i.e. by selecting
them from a scree plot. The default tour layout is a scatter plot with an axis
layout displaying the magnitude and direction of each basis vector. Since the
tour is dynamic, it is often useful to be able to pause and highlight a
particular view. In our interface, brushing will pause the tour path, allowing
users to identify 'interesting' projections. The domain of the axis scales from
running a tour is called the half range, and is computed by rescaling the input
data onto hyperdimensional unit cube. We bind the half range to a mouse wheel
event, allowing a user to pan and zoom on the tour view dynamically. This is
useful for peeling back dense clumps of points to reveal structure.

## Posing Queries: multiple views, multiple contexts

## Making comparsions

# Implementation

The `liminal` software is implemented as an open source R package @baseR. All
visualisations produced by the package were constructed using the `vegawidget`
package, which is an R interface to the Vega-Lite grammar of interactive
graphics. Ensemble graphics for arranging tour and embedding views are
generated using the `shiny` packages. We have made extensive use of the `tourr`
package for constructing tour paths.

## Tours as a streaming data problem

## Interactivity for embedding exploration

# Case Studies

## Case Study 1: Fake trees

## Case Study 2: well-clustered smallish single cell

## Case Study 3: largeish single cell data

## Case Study 4: non-spherical clusters, pdfsense

# Discussion

# References

>>>>

---
title: " Casting multiple shadows: high-dimensional interactive data visualisation with tours and embeddings"
author:
  - name: Stuart Lee
  - name: Di Cook
date: "`r Sys.Date()`"
output: 
  bookdown::html_document2:
    toc: false
    number_sections: no
bibliography: liminal.bib
---

@@

There has been a rapid uptake in the use of non-linear dimensionality reduction (NLDR) methods such as t-distributed stochastic neighbour embedding (t-SNE) in the natural sciences as part of cluster orientation and dimension reduction workflows. The appropriate use of these methods is made difficult by their complex parameterisations and the multitude of decisions required to balance the preservation of
local and global structure in the resulting visualisation. We present a visual analytics framework for the pragmatic usage of NLDR methods by combining them with a technique called the tour. A tour is a sequence of interpolated linear projections of multivariate data onto a lower dimensional space. The sequence is displayed as a dynamic visualisation, allowing a user to see the shadows the high-dimensional data casts in a lower dimensional view. By linking the tour to a view
obtained from an NLDR method, we can preserve global structure and through user interactions like spatial linked brushing observe where
the NLDR view may be misleading. We display several case 
studies from single cell genomics, that shows our approach is useful for
cluster orientation tasks. 
The implementation of our framework is available as an R package called `liminal` available at https://github.com/sa-lee/liminal.

## Introduction

High dimensional data is increasingly prevalent in the natural sciences and
beyond but presents a challenge to the analyst in terms of both data cleaning /
pre-processing and visualisation. Methods to embed data from a high-dimensional
space into a low-dimensional one now form a core step of the data analysis
workflow where they are used to ascertain hidden structure and de-noise data for
downstream analysis (thereby nullifying the 'curse of dimensionality').

@@

Non-linear methods generally perform pre-processing on the high-dimensional data
such as generating a neighborhood graph and perform transformations on the
pre-processed form. We restrict our attention to two methods that are commonly
used in the literature: t-distributed stochastic neighbor embedding
(t-SNE) and uniform manifold alignment and projection (UMAP). The t-SNE
algorithm estimates the pairwise similarity of (Euclidean) distances of points in a high
dimensional space using a Gaussian distribution and then estimates a
configuration in the low dimensional embedding space by modelling similarities
using a t-distribution with 1 degree of freedom. The resulting configuration is
the one that minimizes the Kullback-Leibler divergence between the two
distributions. A recent theoretical contribution by @Linderman2019-dq proved
that t-SNE can recover spherical and well separated cluster shapes, and
proposed new approaches for tuning the optimisation parameters. It is a known
problem that t-SNE can have trouble recovering global structure and that
configurations can be highly dependent on how the algorithm is initialised and
parameterized [@Wattenberg2016-ji; @Kobak2019-lm]. UMAP is a method that is
related to LargeVis [@Tang2016-oz], and like t-SNE acts on the k-nearest
neighbor graph. Its main differences are that it uses a different cost function
(cross entropy) which is optimized using stochastic gradient descent and
defines a different kernel for similarities in the low dimensional space. Due
to it's computational speed it's possible to generate UMAP embeddings in more
than three dimensions.

As part of a visualization workflow, it's important to consider the perception
and interpretation of embedding methods as well. @Sedlmair2013-pn showed that
2D scatter plots were mostly sufficient for detecting class separation, however
they noted that often multiple embeddings were required. For the task of
cluster identification, @Lewis2012-ai showed experimentally that novice users
of non-linear embedding techniques were more likely to consider clusters of
points on a 2d scatter plot to be the result of a spurious embedding compared
to advanced users who were aware of the inner workings of the embedding
algorithm.

A complementary approach for visualizing structure in high dimensional data is
the tour. A tour is a sequence of projections of a high dimensional dataset
onto a low-dimensional orthonormal basis matrix, that is represented as a
dynamic visualization. The sequence of generated bases are interpolated to form
the tour path, allowing a user to explore the subspace of projections. A grand
tour corresponds to choosing new bases at random, and can give an overview of
the structure in the data. Instead of picking projections at random, a guided
tour can be used to generate a sequence 'interesting' projections as quantified
by an index function. Given the dynamic nature of the tour, user interaction is
important for controlling and exploring the visualisation: the tour has been
used previously by @Wickham2015-cx as tool for exploring statistical model fits
and by @Buja1996-fk for exploring factorial experimental designs.

While there has been much work on the algorithmic details of the aforementioned
embedding methods, there has been relatively few tools designed to assist users
to interact with these techniques and assist them in making comparisons between
embeddings and performing the aforementioned cluster orientation tasks. Several
interactive interfaces have been proposed for evaluating or using embedding
techniques: the Sleepwalk interface provides a click and highlight
visualisation for colouring points in an embedding according to their distance
in the original high-dimensional space [@Ovchinnikova2019-gf]. The work by
@Pezzotti2017-cz provides a user guided and modified form of the t-SNE
algorithm, that allows users to modified optimisation parameters in real-time.
Similarly, the embedding projector is a web interface to running UMAP, t-SNE or
PCA live in the browser and provides interactions to color points, and
highlights nearest neighbors [@Smilkov2016-hp].

There is no one-size fits all: finding an appropriate embedding for a given
dataset is a difficult and somewhat poorly defined problem. For non-linear
methods, there are a lot of parameters to explore that can have an effect on
the resulting visualisation and interpretation. Interfaces for evaluating
embeddings require interaction but should also be able to be incorporated into
an analysts workflow. We propose a more pragmatic workflow inspired by
incorporating interactive graphics and tours with embeddings that allows users
to see a global overview of their high dimensional data and assists them with
cluster orientation tasks. This workflow is incorporated into an R package
called `liminal` (Available: https://github.com/sa-lee/liminal).

## Dimensionality Reduction

## Visual Design

@@

### Finding Gestalt: focus and context

To investigate latent structure and the shape of a high dimensional dataset, a
tour can be run without the use of an external embedding. It is often useful to
first run principal components on the input as an initial dimension reduction
step, and then tour a subset of those components instead, i.e. by selecting
them from a scree plot. The default tour layout is a scatter plot with an axis
layout displaying the magnitude and direction of each basis vector. Since the
tour is dynamic, it is often useful to be able to pause and highlight a
particular view. In our interface, brushing will pause the tour path, allowing
users to identify 'interesting' projections. The domain of the axis scales from
running a tour is called the half range, and is computed by rescaling the input
data onto hyper-dimensional unit cube. We bind the half range to a mouse wheel
event, allowing a user to pan and zoom on the tour view dynamically. This is
useful for peeling back dense clumps of points to reveal structure.

### Posing Queries: multiple views, many contexts

We have combined the tour view in a side by side layout with 
a scatter plot view that represents the output of an embedding algorithm. 
These views are linked; analysts can brush regions or
highlight collections of points in either view. Linked highlighting can
be performed when points have been previously labeled according to some
discrete structure, i.e. cluster labels are available. This is achieved
via the analyst clicking on groups in the legend, which causes unselected groupings to have their points become less opaque. Consequently, 
simple linked highlighting can alleviate a known downfall of methods such as 
UMAP or t-SNE: that is distances between clusters are misleading. By
highlighting corresponding clusters in the tour view, the analyst can
see the relationship between clusters, and therefore obtain a more accurate 
representation of the topology of their data.

Simple linked brushing is achieved via mouse-click and drag movements. By
default, when brushing occurs in the tour view, the current projection is 
paused and corresponding points in the embedding view are highlighted. 
Likewise, when brushing occurs in the embedding view, corresponding points
in the tour view are highlighted. In this case, an analyst can use brushing
for manually identifying clusters and verifying cluster locations and shapes:
brushing in the embedding view gives analysts a sense of the shape and 
proximity of cluster in high-dimensional space.

### Making comparisons: revising embeddings

As mentioned previously, when using any DR method, we are assuming
the embedding is representative of the high-dimensional dataset it was
computed from. Defining what it means for embedding to be 'representative` or
'faithful' to high-dimensional data is ill-posed and depends on the underlying
task an analyst is trying to achieve. At the very minimum, we are interested
in distortions and diffusions of the high-dimensional data. Distortions
occur when points that are near each other in the embedding view are 
far from each other in the original dataset. This implies that the embedding 
is not continuous. Diffusions occur when points are far from each other in the embedding view are near in the original data.  Whether, points are near or far 
is reliant on the distance metric used; distortions and diffusions can
be thought of as the preservation of distances or the nearest neighbors
graphs between the high-dimensional space and the embedding space. As distances
can be noisy in high-dimensions, ranks can be used instead as has been 
proposed by ... Identifying distortions and diffusions allows an analyst
to investigate the quality of their embedding and revise them iteratively. We
propose this can be done visually using our side-by-side tour and embedding
views. Some quality checks, such as finding distortions can be identified using simple linked brushing, however we can also interrogate them via spatial
brushes and brush composition.

 Look up nearest neighbours graph from points that lie in a brushing region.

Highlight the corresponding neighbours using colour or transparency in the linked view.

- The $k$ nearest neighbours graph can be pre-computed quickly, for either $X$, $Y$ or both.
- Instead of using the neighbour indices, we could use the neighbour distances instead.
- Composition of multiple brushes could be used to show where there are matches/mismatches between nearest neighbour graphs.

Using a linked neighbourhood brush, we can 
visually investigate the nearest neighbour relationships in the 
high-dimensional space via brushing in the embedding view. The user can select
the number of nearest neighbours directly, and modify the distance metric 
used for determining the neighbours. This allows users to interrogate the stability of clusters generated in the embedding view. Multiple brushes
can be used to pose queries in either the tour view or the embedding view;
interface controls allow these brushes to combine using logical operators
such as 'and', 'or', or 'not'. The use of linked brushing goes beyond
simple color highlights, allowing analysts to get a more holistic
view of the effect of an embedding algorithm.

## Software Infrastructure

We have implemented the above design as an open source R package 
called `liminal` [@baseR]. The package allows analysts to construct
concatenated visualisations, via the
Vega-Lite grammar of interactive graphics (using `vegawidget` package)
and provides an interface for constructing brushes and manipulating tour paths
using the `shiny` and `tourr` packages. `liminal` also provides a stand-alone
interface to the tour, in addition to the linked scatterplot interfaces 
discussed above. Below we display the `liminal` API, and our approach to
generating tour paths and user interactions.

### Tours as a streaming data problem

- don't need to realise the entire sequence
- instead generate new bases according to a fixed frame rate
- allows user interactions to play/pause on the current view

### Linking views via brushes

## Case Studies

### Case Study 1: Exploring tree structured data with tours and t-SNE

(ref:fake-trees) Example from PHATE

```{r fake-trees, echo = FALSE, out.width = "50%", fig.cap="(ref:fake-trees)"}
knitr::include_graphics(here::here("img", "tree.png"))
```

#### Using t-SNE

- there's a lot of parameters to tweak
- emphasises locality, distance between clusters of points can be misleading
- size of clusters can be misleading 
- may require a few different runs to capture topology of data
- if there are clusters it will find them

### Case Study 2: Clustering PBMC 10x single cell RNA-seq data

### Case Study 3: Single cell mouse retina data

## Discussion

## Acknowledgements

## References
